---
title: "Transform Data"
---

## Your Turn 1

```{r Your Turn 1, message=FALSE}
## remove all user defined objects in your workspace
rm(list=ls())

library(tidyverse)  # Provides functions used throughout this session

biochem <- read_csv("biochem_long.csv") #Read in data
```

How can you confirm that you have successfully loaded the data file into Rstudio?

```         
1. The code that imported the data did not yield an error
2. Code that references `biochem` object runs without errors
3. The `biochem` object is present in the environment pane
4. All of the above
```

# filter()

```{r}
# Filters the biochem testing dataset to just rows in which the mrn column is equal to 111900
filter(biochem, mrn == 121444)
```

```{r}
# Filters the biochem testing dataset to just rows in which the last_name column is equal to "maxwell"
filter(biochem, last_name=="maxwell")
```

## Your Turn 2

Use the 3 code blocks below, and `filter()`, to show how you would filter the `biochem` dataset.

A. You are wanting to confirm a reference interval for women over 50 for alkaline phophatase. Find every alkaline phosphatase test for female patients aged 50 and over. The column in which age is found is named `age`, gender is found in `gender` and test is in `test`. Assign the new data frame to the object `biochem_3a`.

```{r Your Turn 3a, eval=FALSE}
#replace the ____ with a logical statement identifying tests for patients greater than 50
biochem_3a_1 <- filter(biochem, age >= 50) 
biochem_3a_2 <- filter(biochem_3a_1, test == "alkaline_phosphatase")
biochem_3a <- filter(biochem_3a_2, gender == "Female")
biochem_3a
```

B. We also would only like to include outpatients? Find all of the alkaline phosphtase testing where the patient type (`pat_type`) is equal to "outpatient". Assign the new data frame to the object `biochem_3b`.

```{r Your Turn 3b, eval=FALSE}
#replace the ____ with a logical statement identifying tests for patients with demo_group equal to "client"
biochem_3b <- filter(biochem_3a, pat_type == "outpatient")
biochem_3b
```

## Filter Bonus Challenge (only if you have time)

We may have some data anomalies in the data set. Find all of the biochemistry testing where the patient class (column `pat_type`) is not entered. In these cases the value will be NA. NA stands for "not available". It is the way that R codes missing values. [Hint: See slide titled "Logical Tests"]

```{r Your Turn 3c, eval=FALSE}
biochem_3c <- filter(biochem, is.na(pat_type))
biochem_3c
```

# mutate

```{r}
# The following code creates a data frame with a new column called c_r_tat_mins which contains the collect to receive TAT quantified in minutes.
mutate(biochem, c_r_tat_mins = col_rec_tat*60)
```

## Your Turn 3

Create a new column using the mutate() function that contains the total TAT. Hint: The total TAT is the sum of `col_rec_tat` and `rec_ver_tat`

```{r Your Turn 6, eval=FALSE}
mutate(biochem, total_tat = 60*(col_rec_tat + rec_ver_tat)) 
```

## Replace a column with `mutate()`

```{r}
mutate(biochem, mrn = as.character(mrn))
```

# Steps and the pipe

```{r Steps and the pipe}
# These 3 lines of code are an example of the how we would accomplish a multi-step analysis based on what we've learned so far. 
tat_outliers <- filter(biochem, pat_type == 'inpatient')
tat_outliers <- mutate(tat_outliers, total_tat = 60*col_rec_tat + 60*rec_ver_tat)
tat_outliers <- filter(tat_outliers, total_tat > 45)

# vs

# This code accomplishes the same task with less code
tat_outliers <- biochem %>% 
	filter(pat_type == 'inpatient') %>% 
	mutate(total_tat = 60*col_rec_tat + 60*rec_ver_tat) %>%   
	filter(total_tat > 60)

```

## Your Turn 4

Use `%>%` in the following chunk to write a sequence of functions that:

1.  Filters to tests from the patient type (`pat_type`) of "cancer center"

2.  Adds a column with the total turnaround time (sum of `col_rec_tat` and `rec_ver_tat`)

3.  Filters to include only the rows with a calculated turnaround time greater than 1 hour

Using "\<-", assign the result to a new variable, call it whatever you want.

```{r your_turn_4, eval=FALSE}
cc_outliers <- biochem %>% 
  filter(pat_type == "cancer center")%>% # filter to the pat_type with the name "cancer center"
  mutate(total_tat = col_rec_tat + rec_ver_tat) %>% # calculate total turnaround time
  filter(total_tat > 1)    # filter only turnaround time greater than 10 hours
cc_outliers
```

## Pipe Bonus Challenge (only if you have time)

Use your mouse to select the name of the new data frame from the list of objects in the Environment tab in the upper-right pane of Rstudio. Answer the following questions:

-   What is the typical TAT for the Cancer Center?
-   Given this data, does it appear that there is a chronic problem in TAT for the Cancer Center, or was this test an outlier?

# What else?

## select()

```{r select}
select(biochem, mrn, last_name)   # select columns to keep
select(biochem, -mrn, -last_name) # select columns to remove 
```

### Bonus Your Turn - select

A. Alter the code to select just the `last_name` column. Assign the result to `biochem_2a` and show the resulting data frame by executing the following line with just the name of the object: `biochem_2a`

```{r your_turn_select_a, eval=FALSE}
# Remove the "_______" and replace it with your answer
biochem_2a <- select(biochem, last_name)
biochem_2a
```

B. If you have time, alter this line of code and see if you can select all columns EXCEPT the `last_name` column.

```{r your_turn_select_b, eval=FALSE}
# Remove the "_______" and replace it with your answer
biochem_2b <- select(biochem, -last_name)
biochem_2b
```

### Select Bonus Challenge (only if you have time)

If you have extra time try the following challenge:

In the `biochem` dataset the turnaround time metrics are labeled `col_rec_tat` and `rec_var_tat`. Your goal is to select just these columns. Using what we learned as well as the dplyr cheatsheet which of the following code snippets could you use to select these columns?

```{r select challenge, eval=FALSE}
select(biochem, contains("rec"))
select(biochem, col_rec_tat, rec_ver_tat)
select(biochem, ends_with("tat"))

```

## Arrange

```{r arrange, eval=FALSE}
arrange(biochem, last_name) #Sort in alphabetical order
arrange(biochem, desc(mrn))  #Sort from largest to smallest MRN
```

### Bonus Your Turn - arrange

The column `value` contains the reported result for each test result.

How might you use arrange() to determine the highest and lowest Sodium result in the dataset? (Hint: Try filtering all `sodium` tests and then using arrange on the `value` column.)

```{r your_turn_arrange, eval=FALSE}
#replace the ____ with the name of the column you would like to arrange the dataframe by.
biochem_sodium <- biochem %>%
  filter(test == "sodium") %>%
  select(value) %>%
  arrange(value) 
head(biochem_sodium)
## head returns the first 6 results in the dataframe
tail(biochem_sodium)
## tail returns the last6 results in the dataframe
```

### Pop Quiz

```{r pop quiz}
toy_data <- data.frame(nums=c(50,1000,3),chars=c("50","1000","3"))
arrange(toy_data,nums)
arrange(toy_data,chars)
```

# Take aways

-   Extract variables with `select()`\
-   Extract cases with `filter()`\
-   Arrange cases, with `arrange()`\
-   Make new variables, with `mutate()`\
-   Connect operations with `%>%`
