---
title: "Statistical Analysis in R"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

## Setup

```{r setup}
## remove all user defined objects in your workspace
rm(list=ls())

library(tidyverse)  # Provides functions used throughout this session

biochem <- read_csv("biochem_wide.csv")
```

## Summarize

### What is the distribution of Alkaline Phosphatase results?

```{r}
ggplot(data = biochem) +
  geom_histogram(aes(x = alkaline_phosphatase), binwidth = 1)
```

### Summarize the total number of of alkaline phosphatase results.

```{r}
biochem %>%
  select(mrn, alkaline_phosphatase) %>%
	summarize(test_count = n()) 

```

## Your Turn 1

Fill in gaps in code below to:

a)  Calculate the median alkaline phosphatase concentration and average calcium concentration.

```{r}
stats_1a <- biochem %>%
	summarize(test_count = n(),
	          mean_alp = median(alkaline_phosphatase),
	          mean_ca = mean(total_calcium))
stats_1a
```

b)  Calculate the interquartile range for alkaline phosphatase and the standard deviation for calcium. The function for interquartile range is "IQR" and for standard deviation is "sd"

```{r}
stats_1b <- biochem %>%
	summarize(test_count = n(),
	          median_alp = median(alkaline_phosphatase),
	          mean_ca = mean(total_calcium),
	          iqr_alp = IQR(alkaline_phosphatase),
	          sd_ca = sd(total_calcium))
stats_1b
```
## Other examples of summarize
```{r}
stats_1c <- biochem %>%
  summarize(test_count = n(),
	          lower95_alp = quantile(alkaline_phosphatase, 0.025),
            upper95_alp = quantile(alkaline_phosphatase, 0.975),
	          var_alp = var(alkaline_phosphatase),
	          cv_alp = 100*sd(alkaline_phosphatase)/mean(alkaline_phosphatase))

```


## Your Turn 2

Consider: How would you calculate the total number of tests performed by patient gender?

## group_by

```{r}
biochem %>%
  select(mrn, gender, pat_type) %>%
  group_by(gender, pat_type)
```

## group_by %\>% summarize

```{r}
biochem %>%  
	select(mrn, gender, alkaline_phosphatase) %>%
  group_by(gender) %>%
  summarize(test_count = n())

```

## Your Turn 3

a)  Calculate the median alkaline phosphatase concentration for Men and Women

```{r, eval=FALSE}
biochem %>%
  group_by(gender) %>%
  summarize(median_alp = median(alkaline_phosphatase))
```

b)  (*Extra*) Calculate the median alkaline phosphatase concentration in Men and Women over 51

```{r, eval=FALSE}
biochem %>%
  group_by(gender) %>%
  filter(age > 51) %>%
  summarize(median_alp = median(alkaline_phosphatase))
```

## What else

### Linear Regression

```{r}
lm.model <- lm(blood_urea_nitrogen ~ creatinine + aspartate_aminotransferase, biochem)
summary(lm.model)

ggplot(biochem, aes(creatinine, blood_urea_nitrogen)) +
  geom_point() +
  xlim(0,3) + 
  ylim(0,50)+
  geom_smooth(method="lm", se=FALSE, color="red", formula = y ~ x)

lm.model$coefficients
confint(lm.model)

```

### Logistic Regression

```{r}
##converting to F and M to 1 and 0
biochem2 <- biochem %>%
  mutate(gender2 = ifelse(gender == "Female", 1, 0)) %>%
  filter(age > 51) %>%
  select(-mrn, -pregnancy_status_at_exam, -age, -collect_dt, -receive_dt, -verify_dt, -pat_type, -payor_groups, -last_name, -col_rec_tat, -rec_ver_tat, -gender)
  

lr.model <- glm(gender2 ~ alkaline_phosphatase + alanine_aminotransferase + aspartate_aminotransferase + gamma_glutamyl_transferase,  	# model formula
               data = biochem2,                           	# dataset
               family = "binomial"                    # type of model
  )
summary(lr.model)
exp(coefficients(lr.model))  					                # odds
```
